---
title: "R Notebook"
output: html_notebook
---
# Set working directory and install and import packages
```{r}
setwd("/Users/tildeidunsloth/Desktop/NLP/NLP_exam")
pacman::p_load(tidyverse,brms, posterior, tidybayes)
```

# Fixing up some stan stuff
```{r}
cmdstanr::cmdstan_path()
cmdstanr::cmdstan_version()
```


# Avoid scientific notation
```{r}
options(scipen = 999)
```


# Import the data
```{r}
# Loading in the data
data <- read_csv("word_similarity__extra_categories_results.csv")

# Changing the year variable to make summary output easier to interpret
data <- data %>% 
  mutate(year_centered = year - 1900)
```
# Subsets for different categories
```{r}
# Making subsets of data based on category
positive_agency <- data %>% 
  filter(category == "positive_agency")
negative_agency <- data %>% 
  filter(category == "negative_agency")
high_competence <- data %>% 
  filter(category == "high_competence")
low_competence <- data %>% 
  filter(category == "low_competence")
high_warmth <- data %>% 
  filter(category == "high_warmth")
low_warmth <- data %>% 
  filter(category == "low_warmth")
occupation_trad_female <- data %>% 
  filter(category == "occupation_trad_female")
occupation_trad_male <- data %>% 
  filter(category == "occupation_trad_male")
appearance_related <- data %>% 
  filter(category == "appearance_related")
```

# Investigating the distribution of values for the variable we want to predict
```{r}
# Looking at the distribution
hist(data$gender_bias, breaks = 30)
```


# Plotting gender bias over time for specific words
```{r}
ggplot(data[data$word == "friendly", ], aes(x = year_centered, y = gender_bias)) +
  geom_line(stat = "summary", fun = "mean", color = "blue") +
  geom_point(stat = "summary", fun = "mean", color = "red") +
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  labs(title = "Gender Bias Over Time for word friendly",
       x = "Year",
       y = "Gender Bias") +
  theme_minimal()

ggplot(data[data$word == "independent", ], aes(x = year_centered, y = gender_bias)) +
  geom_line(stat = "summary", fun = "mean", color = "blue") +
  geom_point(stat = "summary", fun = "mean", color = "red") +
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  labs(title = "Gender Bias Over Time for word independent",
       x = "Year",
       y = "Gender Bias") +
  theme_minimal()

ggplot(data[data$word == "enforce", ], aes(x = year_centered, y = gender_bias)) +
  geom_line(stat = "summary", fun = "mean", color = "blue") +
  geom_point(stat = "summary", fun = "mean", color = "red") +
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  labs(title = "Gender Bias Over Time for word enforce",
       x = "Year",
       y = "Gender Bias") +
  theme_minimal()

ggplot(data[data$word == "nurse", ], aes(x = year_centered, y = gender_bias)) +
  geom_line(stat = "summary", fun = "mean", color = "blue") +
  geom_point(stat = "summary", fun = "mean", color = "red") +
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  labs(title = "Gender Bias Over Time for word nurse",
       x = "Year",
       y = "Gender Bias") +
  theme_minimal()

ggplot(data[data$word == "command", ], aes(x = year_centered, y = gender_bias)) +
  geom_line(stat = "summary", fun = "mean", color = "blue") +
  geom_point(stat = "summary", fun = "mean", color = "red") +
  geom_smooth(method = "lm", color = "green", se = TRUE) +
  labs(title = "Gender Bias Over Time for word command",
       x = "Year",
       y = "Gender Bias") +
  theme_minimal()
```



# Setting a broad prior for the bayesian models
```{r}
# Setting the prior
prior <- c(
  prior(normal(0, 0.1), class = "b"),        # Slope: small changes over time
  prior(normal(0, 1), class = "Intercept"),  # Intercept: centered around observed values
  prior(exponential(1), class = "sigma")     # Residual SD: positive and moderate spread
)

```

# Looking at the values and variance for the variables of interest
```{r}
summary(data$gender_bias)
var(data$gender_bias)
```

# Positive agency model
```{r}
cores <- parallel::detectCores()
chains = 2

# Trying to run a model
model_pos_agency <- brm(
  formula = gender_bias ~ year_centered,
  data = positive_agency,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)

```

```{r}
print(summary(model_pos_agency), digits = 5)
plot(model_pos_agency)

# View the summary of the model
posterior_summary(model_pos_agency)

pp_check(model_pos_agency)
```


```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_pos_agency %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(positive_agency$year_centered), 
                         max(positive_agency$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Save the plot as a ggplot object
plot_positive_agency <- ggplot(positive_agency, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) +
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),
    .width = c(0.5, 0.8, 0.95),
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Positive Agency",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()

```



# Negative agency model
```{r}
model_negative_agency <- brm(
  formula = gender_bias ~ year_centered,
  data = negative_agency,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```



```{r}
print(summary(model_negative_agency), digits = 5)
plot(model_negative_agency)

pp_check(model_negative_agency)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_negative_agency %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(negative_agency$year_centered), 
                         max(negative_agency$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_negative_agency <- ggplot(negative_agency, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Negative Agency",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()

```

# High Warmth model
```{r}
# Trying to run a model
model_high_warmth <- brm(
  formula = gender_bias ~ year_centered,
  data = high_warmth,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```
```{r}
print(summary(model_high_warmth), digits = 5)
plot(model_high_warmth)
pp_check(model_high_warmth)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_high_warmth %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(high_warmth$year_centered), 
                         max(high_warmth$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_high_warmth <- ggplot(high_warmth, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "High Warmth",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```


# Low warmth model
```{r}
# Trying to run a model
model_low_warmth <- brm(
  formula = gender_bias ~ year_centered,
  data = low_warmth,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_low_warmth), digits = 5)
plot(model_low_warmth)
pp_check(model_low_warmth)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_low_warmth %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(low_warmth$year_centered), 
                         max(low_warmth$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_low_warmth <-ggplot(low_warmth, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Low Warmth",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```


# Low competence model
```{r}
model_low_competence <- brm(
  formula = gender_bias ~ year_centered,
  data = low_competence,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_low_competence), digits = 5)
plot(model_low_competence)
pp_check(model_low_competence)
```
```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_low_competence %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(low_competence$year_centered), 
                         max(low_competence$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_low_competence <-ggplot(low_competence, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Low Competence",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```



# High competence model
```{r}
model_high_competence <- brm(
  formula = gender_bias ~ year_centered,
  data = high_competence,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```


```{r}
print(summary(model_high_competence), digits = 5)
plot(model_high_competence)
pp_check(model_high_competence)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_high_competence %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(high_competence$year_centered), 
                         max(high_competence$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_high_competence <- ggplot(high_competence, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "High Competence",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```


# Traditional female occupation model
```{r}
model_trad_female <- brm(
  formula = gender_bias ~ year_centered,
  data = occupation_trad_female,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_trad_female), digits = 5)
plot(model_trad_female)
pp_check(model_trad_female)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_trad_female %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(occupation_trad_female$year_centered), 
                         max(occupation_trad_female$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_trad_female <- ggplot(occupation_trad_female, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Traditional female jobs",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```


# Traditional male occupation model
```{r}
model_trad_male <- brm(
  formula = gender_bias ~ year_centered,
  data = occupation_trad_male,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_trad_male), digits = 5)
plot(model_trad_male)
pp_check(model_trad_male)
```
```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_trad_male %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(occupation_trad_male$year_centered), 
                         max(occupation_trad_male$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_trad_male <- ggplot(occupation_trad_male, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Tradtional male jobs",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```



# Appearance related words
```{r}
model_appearance_related <- brm(
  formula = gender_bias ~ year_centered,
  data = appearance_related,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_appearance_related), digits = 5)
plot(model_appearance_related)
pp_check(model_appearance_related)
```
```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_appearance_related %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(appearance_related$year_centered), 
                         max(appearance_related$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_appearance_related <- ggplot(appearance_related, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "Appearance related words",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()
```


# Combined mega plot
```{r}
# Remove the legend (sidebar) from each plot
plot_positive_agency <- plot_positive_agency + theme(legend.position = "none")
plot_negative_agency <- plot_negative_agency + theme(legend.position = "none")
plot_high_competence <- plot_high_competence + theme(legend.position = "none")
plot_low_competence <- plot_low_competence + theme(legend.position = "none")
plot_low_warmth <- plot_low_warmth + theme(legend.position = "none")
plot_high_warmth <- plot_high_warmth + theme(legend.position = "none")
plot_trad_female <- plot_trad_female + theme(legend.position = "none")
plot_trad_male <- plot_trad_male + theme(legend.position = "none")
plot_appearance_related <- plot_appearance_related + theme(legend.position = "none")

# Load the patchwork library
library(patchwork)

# Combine plots into a 2-column layout with a blank spacer
combined_plot <- (plot_positive_agency + plot_negative_agency + 
                  plot_high_competence + plot_low_competence +
                  plot_high_warmth + plot_low_warmth +
                  plot_trad_female + plot_trad_male +
                  plot_appearance_related + plot_spacer()) +
  plot_layout(ncol = 2, heights = c(1, 1, 1, 1, 1))  # Define layout with 2 columns and row heights

# View the combined plot
combined_plot

# Save the plot with proper dimensions
ggsave("combined_plots_same_width.png", plot = combined_plot, width = 12, height = 18, dpi = 300)



```

# Gender bias for all categories combined
```{r}
model_all <- brm(
  formula = gender_bias ~ year_centered,
  data = data,
  backend = "cmdstanr",
  prior = prior,
  family = gaussian(),
  chains = chains,
  iter = 3000,
  warmup = 1000,
  cores = cores
)
```

```{r}
print(summary(model_all), digits = 5)
plot(model_all)
pp_check(model_all)
```

```{r}
# Step 1: Extract posterior draws for the regression line
posterior_lines <- model_all %>%
  spread_draws(b_Intercept, b_year_centered) %>%            # Extract posterior samples
  mutate(year = list(seq(min(data$year_centered), 
                         max(data$year_centered), length.out = 100))) %>%
  unnest(year) %>%
  mutate(.fitted = b_Intercept + b_year_centered * year)    # Compute fitted values for each 'year'

# Step 2: Plot the observed data and posterior regression
plot_all <- ggplot(data, aes(x = year_centered, y = gender_bias)) +
  geom_point(aes(color = word), alpha = 0.7, size = 2) + # Observed data (one word/year)
  stat_lineribbon(
    data = posterior_lines,
    aes(x = year, y = .fitted),                    # Posterior fitted line
    .width = c(0.5, 0.8, 0.95),                    # Add credible intervals
    fill = "lightblue", alpha = 0.5
  ) +
  ylim(-0.2, 0.2) +
  labs(title = "All Categories combined",
       x = "Year",
       y = "Gender Bias",
       color = "Word") +
  theme_minimal()

plot_all
```

